---
title: "R Workshop: ANOVA, ANCOVA & Factorial ANOVA"
author: "Brier Gallihugh, M.S."
date: today
format:
  html:
    theme: default
    code-annotations: select
    self-contained: true
toc: true
toc-location: right
toc-depth: 3
warning: FALSE
echo: true
---

## Introduction to ANOVA

## Running ANOVA

```{r running anova}
library(car)
library(psych)
library(multcomp)
library(effects)
library(tidyverse)
library(sjstats)


data <- starwars

data <- data %>% 
  dplyr::select(height,mass,hair_color,species,sex) %>% 
  na.omit() %>% 
  filter(species == "Gungan" | species == "Human" | species == "Wookiee")
  
height_species_aov <- aov(height ~ species, data = data)
summary(height_species_aov)

# Post Hoc Test (Tukey)
TukeyHSD(height_species_aov)
# Bonferroni Adjustment vs Tukey
pairwise.t.test(data$height, data$species, p.adjust.method = "bonferroni")
```

## Assumptions of ANOVA

### DV Normality

#### Graphical

```{r anova normality graph}
density_plot <- data %>% ggplot(aes(x = height)) +
  geom_histogram(aes(y= after_stat(density)),binwidth = 1) +
  stat_function(fun = dnorm,
                args = list(mean = mean(data$height),
                            sd = sd(data$height)),
                col = "blue",
                linewidth = 1) +
  labs(title = "Figure 1. Histogram of Height Scores",
       x = "Height Scores",
       y = "Density")

density_plot

# OR

qqnorm(data$height)
qqline(data$height, col = "blue")
```

#### Statistical

```{r anova normality stat}
psych::describe(data$height)
shapiro.test(data$height)
```

### Homogeneity of Variance

#### Graphical

```{r anova homogeneity graph}
boxplot(data$height ~ data$species)
```

#### Statistical

```{r anova homogeneity stat}
leveneTest(data$height ~ data$species)
```

## Introduction to ANCOVA

## Running ANCOVA

```{r running ancova}
data <- data %>% mutate(species = as.factor(species))
# Type I SS
ancova <- aov(height ~ species + mass, data = data)
summary(ancova)
# Reversed Order
ancova2 <- aov(height ~ mass + species, data = data)
summary(ancova2)
# Values are DIFFERENT. Why? R by default does ANOVA sequentially (which you usually don't want)

# A fix is to use Type III SS. We can do that using the following code
Anova(ancova, type = "III")

# Std Means
summary_data <- data %>% 
  dplyr::group_by(species) %>% 
  dplyr::summarise(n = n(),
                   mean = mean(height))

# Adjust means for covariate effect
adjustedMeans<- effect("species", ancova, se=TRUE)
summary(adjustedMeans)
adjustedMeans$se

# Post Hoc Tests
posthoc <- glht(ancova, linfct = mcp(species = "Tukey"))
summary(posthoc)
confint(posthoc)

# Effect size
sjstats::anova_stats(Anova(ancova,type = "III"))
```

## Assumptions of ANCOVA

### Predictor x Covariate Indepenence

#### Statistical

```{r predictor indep anova}
predictor_assumption <- aov(mass ~ species, data = data)
summary(predictor_assumption)

# Here it's statistically significant (which is bad). It actually means we can't do this (unless we have random assignment) but we will anyway for the sake of an example
```

### Homogeneity of Regression Slopes

#### Statistical

```{r homogeneity of regression slopes stat}
regression_slope_assumption <- aov(height ~ species*mass, data = data)
Anova(regression_slope_assumption, type = "III")

# Interaction in NOT significant so assumption is met
```

### Residual Normality (DV ~ IV)

#### Statistical

```{r ancova residual normality stat}
height_mass_resid <- scale(residuals(aov(height ~ mass, data = data)))
describe(height_mass_resid)

shapiro.test(height_mass_resid)
```

#### Graphical

```{r ancova residual normality graph}
hist(height_mass_resid, col = 'beige', 
     main="", xlab = "ANCOVA Residuals (z-score)",
     probability = TRUE)
curve(dnorm(x, mean = mean(height_mass_resid), 
            sd = sd(height_mass_resid)), 
            add = TRUE, lwd = 2, col = 'blue')
```

### Homogeneity of Variance

#### Statistical

```{r homogeneity stat}
# Levene's Test to Assess Equal Variance for Species
leveneTest(data$height ~ data$species)
```

#### Graphical

```{r homogeneity graph}
# Boxplot Height by Species
boxplot(height ~ species,data=data, main="Height Variance by Species ",
   xlab="Species", ylab="Height") 
```

### Linearity of CV & DV

#### Graphical

```{r ancova linearity graph}
plot(data$height ~ data$mass,
     pch = 16, bty = 'l',
     xlab = "Mass",
     ylab = "Height")
```

## Introduction to Factorial ANOVA

## Running Factorial ANOVA

```{r running factorial anova}
factorial_data <- starwars %>%
  select(mass,homeworld,height) %>% 
  mutate(homeworld = as.factor(homeworld)) %>%
  filter(homeworld == "Tatooine" | homeworld == "Naboo") %>% 
  mutate(homeworld = dplyr::recode(homeworld,
                                   "Tatooine" = 0,
                                   "Naboo" = 1))

factorial_data <- factorial_data %>% 
  mutate(homeworldMC = factorial_data$homeworld - mean(factorial_data$homeworld),
         heightMC = factorial_data$height - mean(factorial_data$height))

# Modeling As A Regression (For SS Analyses)
reg_fanova <- lm(mass ~ heightMC*homeworldMC,data = factorial_data)
summary(reg_fanova)

# Modeling As Factorial ANOVA
aov_factorial <- aov(mass ~ height*homeworld, data = factorial_data)
Anova(aov_factorial, type = "III")
```

```{r simple slopes factorial anaova}
library(pequod)

fit <- pequod::lmres(mass ~ heightMC*homeworldMC,data = factorial_data)
slope_homeworld <- pequod::simpleSlope(fit,pred = "heightMC", mod1 = "homeworldMC")
summary(slope_homeworld)
PlotSlope(slope_homeworld)

slope_height <- pequod::simpleSlope(fit,pred = "homeworldMC", mod1 = "heightMC")
summary(slope_height)
PlotSlope(slope_height)
```

## Assumptions of Factorial ANOVA

### Linearity of IV to DV (Regression)

#### Graphical

```{r normality residuals graph}
plot(lm(mass ~ height, data = factorial_data),2)

plot(lm(mass ~ homeworld, data = factorial_data),2)
```

#### Statistical

```{r normality residuals stat}
psych::describe(resid(aov_factorial))
shapiro.test(resid(aov_factorial))
```

### Homogeneity of Variance (Regression)

#### Graphical

```{r homogeneity of variance fanova graph}
boxplot(factorial_data$mass ~ factorial_data$height)

boxplot(factorial_data$mass ~ factorial_data$homeworld)

# Residuals
f_anova_residuals <- scale(residuals(lm(mass ~ heightMC*homeworldMC, data = factorial_data)))
f_anova_fitted <- fitted(lm(mass ~ heightMC*homeworldMC, data = factorial_data))

plot(f_anova_residuals ~ f_anova_fitted)
```

#### Statistical

```{r homogeneity variance fanova stat}
leveneTest(factorial_data$mass ~ factorial_data$height)

leveneTest(factorial_data$mass ~ factorial_data$homeworld)

leveneTest(factorial_data$mass ~ interaction(factorial_data$homeworld,factorial_data$height))
```

### Normallity of Residuals

#### Graphical

```{r normality of residuals factorial graph}
plot(aov_factorial,2)
```

#### Statistical

```{r normality residuals factorial stat}
fanova_residuals <- aov_factorial$residuals

shapiro.test(fanova_residuals)

psych::describe(fanova_residuals)
```
