---
title: "R Workshop: Mediation and Moderation"
author: "Brier Gallihugh, M.S."
date: today
format:
  html:
    theme: default
    code-annotations: select
    self-contained: true
toc: true
toc-location: right
toc-depth: 3
warning: FALSE
echo: true
---

```{r library}
set.seed(10311993)
library(mediation)
library(psych)
library(tidyverse)

# Created Toy Data Set
# Variance Covariance
sigma <- rbind(c(1,-0.4,-0.3), c(-0.4,1, 0.7), c(-0.3,0.7,1))
# Variable Mean
mu <- c(7, 50, 7) 
# Generate the Multivariate Normal Distribution
df <- as.data.frame(mvrnorm(n=100, mu=mu, Sigma=sigma))
df <- round(df,0)
colnames(df) <- c("mediator1","outcome","predictor")
df$condition <- rep(1:2,50)
```

## Running a Moderation Analysis in R

```{r running moderation}
moderation <- lm(outcome ~ condition*predictor, data = df)
summary(moderation)
```

## Running a Mediation Analysis in R

```{r running mediation}
#Regress M on X
outcomeM_fit <- lm(mediator1 ~ condition, data = df)
summary(outcomeM_fit)

#Regress Y on M and X
outcomeY_fit <- lm(outcome ~ mediator1 + condition, data = df)
summary(outcomeY_fit)

#Run Mediation with Bootstrap
outcome_fit <- mediation::mediate(outcomeM_fit,
                                  outcomeY_fit,
                                  treat = "condition",
                                  mediator = "mediator1",
                                  boot = TRUE,
                                  sims = 5000)
#Summary of Mediation
summary(outcome_fit)

#Path Coefficients
plot(outcome_fit)
```

## Assumptions of Moderation Analyses

```{r moderation assumptions}
# Residual Normality
shapiro.test(residuals(moderation))

# Multicollinearity
car::vif(moderation, type = c("predictor"))

# Independence of Errors
car::durbinWatsonTest(moderation)
```

## Assumptions of Mediation Analyses

```{r mediation assumptions}
# Linearity
plot(lm(outcome ~ predictor, data = df),2)
plot(lm(outcome ~ mediator1, data = df),2)
plot(lm(mediator1 ~ predictor, data = df),2)
# Multicollinearity
cor.plot(df)
```

# Using Moderation and Mediation Usings Hayes PROCESS Macro (for R)

Click on the following [link](https://www.afhayes.com/public/processv43.zip) to download the R script for the PROCESS macro for R.

```{r run hayes file}
source("process.R")
```

## A Moderation Example Using Hayes PROCESS Macro

```{r running moderation with hayes}
process(data = df,
        y = "outcome",
        x = "predictor",
        w = "mediator1",
        model = 1,
        stand = 1)

```

## A Mediation Example Using Hayes PROCESS Macro

```{r running mediation with hayes}
process(data = df,
        y = "outcome",
        x = "predictor",
        m = "mediator1",
        model = 4,
        stand = 1,
        boot = 5000,
        seed = 10311993)
```
